Association between trauma severity models and opportunities for improvement: A retrospective cohort study

### 2022-09-11 project-timeline.rmd
Comments JA 22.09.11

# Project timeline

## Week 1 (29/9 - 2/9)
- Setup Github ([done](https://github.com/victoria-l-w/trauma-severity-ofi))
- Draft study progress report
- Reading

>>>>

# Project timeline
<!--Mycket mycket bra Victoria! -->
## Week 1 (29/9 - 2/9)
- Setup Github ([done](https://github.com/victoria-l-w/trauma-severity-ofi))
- Draft study progress report
- Reading

### 2022-09-11 studyprogressreport.Rmd
Comments JA 22.09.11

Introduction
============

<!-- Why trauma is a big deal -->

@@

The Trauma and Injury Severity Score (TRISS) is one of the most commonly used models. TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss]. TRISS estimates the possibility of survival for the given parameters, and can be used by hospitals as a benchmark to compare outcomes against a standardised survival estimate.

<!-- Info on other scores here? -->

<!-- Explain the what and why of M&M -->

Another cornerstone of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality conferences (MMC). Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@errors].  These conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case and to implement corrective actions. OFI function in this manner as a proxy for suboptimal care. The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

TSS are already used in some clinics to identify cases for peer review [@whoguidelines]. However, the extent to which TSS are associated with or can predict opportunities for improvement (OFI) as found in M&M conferences is poorly understood. If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.

@@

The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021. This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines. The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment. Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.

@@

Participants include all patients registered in both the trauma registry and the trauma care quality registry at KUH. The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated. Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15. Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.

<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at NKS? -->

The trauma care quality registry contains patients from the trauma registry that were selected for M&M review. At KUH, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are evaluated by research nursing staff to identify cases that potentially received sub-optimal care. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.

@@

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
The exposure variables are the values given by the studied trauma models. These variables are continuous. The outcome variable, OFI, is a binary variable.

<!-- Really not sure exactly what to put here
-->

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

<!-- Complete case study? -->
<!--

>>>>

Introduction
============
<!--Mycket bra! Röd tråd och "tratt". -->
<!-- Why trauma is a big deal -->

@@

The Trauma and Injury Severity Score (TRISS) is one of the most commonly used models. TRISS is a combination index that incorporates patient age, injury severity by anatomical site, and physiological parameters [@triss]. TRISS estimates the possibility of survival for the given parameters, and can be used by hospitals as a benchmark to compare outcomes against a standardised survival estimate. <!--Jag skulle nämna dina andra scores här -->

<!-- Info on other scores here? -->
<!-- Explain the what and why of M&M --> 
<!-- -->
Another cornerstone of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality conferences (MMC). Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@errors].  These conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case and to implement corrective actions. OFI function in this manner as a proxy for suboptimal care. The WHO guidelines for trauma quality improvement programmes include specific recommendations on holding regularly scheduled M&M conferences, but also note that M&M conferences are not used to their fullest potential globally [@whoguidelines]. <!--Kolla exv Strömmers och Ghorbanis pek,https://pubmed-ncbi-nlm-nih-gov.sll.idm.oclc.org/29797712/ .
De försöker använda bl.aa normit och triss för att se korrelans till OFI och preventable deaths. Skillnaden mot det du gör är att de bara tittar på patienter som dör och inte hela cohorten. Du tittar också på ett betydligt större dataunderlag. Så ditt bygger vidare på deras arbete-->

@@

TSS are already used in some clinics to identify cases for peer review [@whoguidelines]. However, the extent to which TSS are associated with or can predict opportunities for improvement (OFI) as found in M&M conferences is poorly understood. If it is found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. This study aims to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.
<!--Introduktionen är mycket bra! Egentligen bara fortsätta som du gör, läs på och lägg till meningar/källor när du hittar info. För att nå KI:s sidkrav kan du skriva en hel del om de olika Scorsen -->

@@

The trauma registry at KUH contains data on 21,000 patients collected between 2012 and 2021.<!--Det kanske är så många? Var hittade du den siffran? --> This registry is also submitted to SweTrau, Sweden's national trauma registry database, and therefore meets SweTrau guidelines. The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment. Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units. <!--Jag skulle nämna att trauma registret följer "the utstein template" vilket är en internationell konsensus kring hur register byggs, inkl trauma register i detta fall -->

@@

Participants include all patients registered in both the trauma registry and the trauma care quality registry at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här --> The trauma registry includes all patients for whom the trauma team was activated after receiving a potentially traumatic injury, regardless of NISS score, and all patients with a NISS score >15, regardless of whether the trauma team was activated. Also included are patients who were transferred to the hospital within 7 days of injury and have a NISS score >15. Patients where the only traumatic injury is a chronic subdural hematoma and patients for whom the trauma team was activated without an underlying traumatic injury are excluded from the register.

<!-- Eligiblity criteria for the trauma registry were taken from the SWETRAU manual, check if it is exactly the same at NKS? -->
<!--Det stämmer! -->

The trauma care quality registry contains patients from the trauma registry that were selected for M&M review. At KUH, all fatalities are automatically selected for M&M review. In addition to this, the electronic medical records of all trauma patients are evaluated by research nursing staff to identify cases that potentially received sub-optimal care. Cases that two research nurses have evaluated and found to have potential for sub-optimal care are also selected for M&M review.
<!--Helt korrekt, de använder även "audit filter" i sin screening, ev behöver du nämna detta också, se tid studenters arb. -->

@@

Quantitative variables
----------------------
<!--Om du behöver "skapa" ny data genom att manipulera den insamlade (tvek i ditt fall) så ska du beskriva det här. Annars kan detta stycke slås ihop med variables ovan. -->
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
The exposure variables are the values given by the studied trauma models. These variables are continuous. The outcome variable, OFI, is a binary variable. 
<!--Om du inte tydligt definerat modellernas variabler behöver du göra det här, alltså för exv triss som är GCS, RR, BT + ISS osv osv-->
<!-- Really not sure exactly what to put here
-->

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant. 
<!-- Jo det kommer nog bli ccs, du kommer exkludera alla patienter som delsa saknar outcome, OFI alternativt där det saknas nog data för att bygga olika scores. -->
<!-- Complete case study? -->
<!--
